The Role of Variability in Learning Generalization: A Computational Modeling Approach

### 2024-06-22 Sections/full.qmd
Rob's edits up through end of htw into

::: {.content-visible when-format="pdf"}

{{< include /Sections/pre_sections/toc_pdf.qmd >}}

:::

@@

In addition to largely overlooking the potential for non-linear generalization to confound interpretations of training manipulations, the visuomotor skill learning literature also rarely considers alternatives to schema representations [@chamberlinMemoryRepresentationMotor1992]. Although schema-theory remains influential within certain literatures, instance or exemplar-based models have accounted for human behavior across myriad domains [@jamiesonInstanceTheoryDomaingeneral2022; @loganInstanceTheoryAttention2002a]. As mentioned above, instance based accounts have been shown to perform well on a variety of different tasks with motoric components [@crumpEpisodicContributionsSequential2010;@gandolfoMotorLearningField1996a;  @meighWhatMemoryRepresentation2018; @rosenbaumPlanningReachesEvaluating1995; @vandamMappingShapeVisuomotor2015]. However, such accounts have received little attention within the subdomain of visuomotor skill learning focused on the benefits of varied training.

@@

::: {#tbl-brms-models}
| Group Comparison         | Code                                                        | Data                 |
| ------------------------ | ----------------------------------------------------------- | -------------------- |
| End of Training Accuracy | `` brm(Abs. Deviation ~ condit) ``                                    | Final Training Block |
| Test Accuracy            | `` brm(Abs. Deviation ~ condit * bandType + (1|id) + (1|bandInt) `` | All Testing trials   |
| Band Discrimination      | `` brm(vx ~ condit * band +(1 + bandInt|id) ``             | All Testing Trials   |

: **Statistical Model Specifications**. The specifications for the Bayesian regression models used in the analyses of each of the 3 experiments. Comparisons of accuracy use abosulte deviation as the dependent variable, while comparisons of discrimination use the raw velocities produced by participants as the dependent variable.
:::
\

@@

::: {#tbl-e1-train-dist}

| Term         | Estimate | 95% CrI Lower | 95% CrI Upper |  pd |
|:-------------|---------:|--------------:|--------------:|----:|
| Intercept    |   106.34 |         95.46 |        117.25 |   1 |
| conditVaried |    79.64 |         57.92 |        101.63 |   1 |

: **Experiment 1 - End of training performance**. Comparing final training block accuracy in band common to both groups. The Intercept represents the average of the baseline condition (constant training), and the conditVaried coefficient reflects the difference between the constant and varied groups. A larger positive estimates indicates a greater deviation (lower accuracy) for the varied group. CrI values indicate 95% credible intervals. pd is the probability of direction (the % of the posterior on the same side of 0 as the coefficient estimate). {.sm}

:::

@@

::: {#tbl-e1-bmm-dist}

| Term                               | Estimate | 95% CrI Lower | 95% CrI Upper |  pd |
|:------------------------|-----------:|-----------:|-----------:|-----------:|
| Intercept                          |   152.55 |         70.63 |        229.85 | 1.0 |
| conditVaried                       |    39.00 |        -21.10 |        100.81 | 0.9 |
| bandTypeExtrapolation              |    71.51 |         33.24 |        109.60 | 1.0 |
| conditVaried:bandTypeExtrapolation |    66.46 |         32.76 |         99.36 | 1.0 |

@@

::: {#tbl-e1-bmm-vx}

| Term         | Estimate | 95% CrI Lower | 95% CrI Upper |   pd |
|:-------------|---------:|--------------:|--------------:|-----:|
| Intercept    |   408.55 |        327.00 |        490.61 | 1.00 |
| conditVaried |   164.05 |         45.50 |        278.85 | 1.00 |
| Band         |     0.71 |          0.62 |          0.80 | 1.00 |
| condit\*Band |    -0.14 |         -0.26 |         -0.01 | 0.98 |

@@

::: {#tbl-e2-train-dist}

| Term         | Estimate | 95% CrI Lower | 95% CrI Upper |  pd |
|:-------------|---------:|--------------:|--------------:|----:|
| Intercept    |    91.01 |         80.67 |        101.26 |   1 |
| conditVaried |    36.15 |         16.35 |         55.67 |   1 |

@@

*Training*. @fig-e2-train-dev presents the deviations across training blocks for both constant and varied training groups. We again compared training performance on the band common to both groups (600-800). The full model results are shown in Table 1. The varied group had a significantly greater deviation than the constant group in the final training block, ( $\beta$ = `r cdtr2$estimate`, 95% CrI \[`r cdtr2$conf.low`, `r cdtr2$conf.high`\]; pd = `r cdtr2$pd`).
```{r}
modelFile <- paste0(here::here("data/model_cache/"), "e2_dist_Cond_Type_RF_2")
bmtd2 <- brm(dist ~ condit * bandType + (1|bandInt) + (1|id), 
    data=testE2, file=modelFile,
    iter=5000,chains=4, control = list(adapt_delta = .94, max_treedepth = 13))

@@

::: {#tbl-e2-bmm-dist}

| Term                               | Estimate | 95% CrI Lower | 95% CrI Upper |   pd |
|:-----------------------------------|---------:|--------------:|--------------:|-----:|
| Intercept                          |   190.91 |        125.03 |        259.31 | 1.00 |
| conditVaried                       |   -20.58 |        -72.94 |         33.08 | 0.78 |
| bandTypeExtrapolation              |    38.09 |         -6.94 |         83.63 | 0.95 |
| conditVaried:bandTypeExtrapolation |    82.00 |         41.89 |        121.31 | 1.00 |

@@

::: {#tbl-e2-bmm-vx}

| Term         | Estimate | 95% CrI Lower | 95% CrI Upper |   pd |
|:-------------|---------:|--------------:|--------------:|-----:|
| Intercept    |   362.64 |        274.85 |        450.02 | 1.00 |
| conditVaried |    -8.56 |       -133.97 |        113.98 | 0.55 |
| Band         |     0.71 |          0.58 |          0.84 | 1.00 |
| condit\*Band |    -0.06 |         -0.24 |          0.13 | 0.73 |

@@

::: {#tbl-e3-train-dist}

| Term                          | Estimate | 95% CrI Lower | 95% CrI Upper |   pd |
|:------------------------------|---------:|--------------:|--------------:|-----:|
| Intercept                     |   121.86 |        109.24 |        134.60 | 1.00 |
| conditVaried                  |    64.93 |         36.99 |         90.80 | 1.00 |
| bandOrderReverse              |     1.11 |        -16.02 |         18.16 | 0.55 |
| conditVaried:bandOrderReverse |   -77.02 |       -114.16 |        -39.61 | 1.00 |

: **Experiment 3 - End of training performance**. The Intercept represents the average of the baseline condition (constant training & original band order), the conditVaried coefficient reflects the difference between the constant and varied groups, and the bandOrderReverse coefficient reflects the difference between original and reverse order. A larger positive coefficient indicates a greater deviation (lower accuracy) for the varied group. The negative value for the interaction between condit and bandOrder indicates that varied condition with reverse order had significantly lower deviations than the varied condition with the original band order  {.sm}

:::

@@

::: {#tbl-e3-bmm-dist}
| Term                                                | Estimate | 95% CrI Lower | 95% CrI Upper |   pd |
|:----------------------------------------------------|---------:|--------------:|--------------:|-----:|
| Intercept                                           |   288.65 |        199.45 |        374.07 | 1.00 |
| conditVaried                                        |   -40.19 |       -104.68 |         23.13 | 0.89 |
| bandTypeExtrapolation                               |   -23.35 |        -57.28 |         10.35 | 0.92 |
| bandOrderReverse                                    |   -73.72 |       -136.69 |        -11.07 | 0.99 |
| **conditVaried:bandTypeExtrapolation**                  |    52.66 |         14.16 |         90.23 | 1.00 |
| conditVaried:bandOrderReverse                       |   -37.48 |       -123.28 |         49.37 | 0.80 |
| bandTypeExtrapolation:bandOrderReverse              |    80.69 |         30.01 |        130.93 | 1.00 |
| conditVaried:bandTypeExtrapolation:bandOrder |    30.42 |        -21.00 |         81.65 | 0.87 |

@@

::: {#tbl-e3-bmm-vx}

| Term                                    | Estimate | 95% CrI Lower | 95% CrI Upper |   pd |
|:----------------------------------------|---------:|--------------:|--------------:|-----:|
| Intercept                             |   601.83 |        504.75 |        699.42 | 1.00 |
| conditVaried                          |    12.18 |       -134.94 |        162.78 | 0.56 |
| bandOrderReverse                      |    13.03 |       -123.89 |        144.67 | 0.58 |
| **Band**                                  |     0.49 |          0.36 |          0.62 | 1.00 |
| **conditVaried:bandOrderReverse**         |  -338.15 |       -541.44 |       -132.58 | 1.00 |
| conditVaried:Band                     |    -0.04 |         -0.23 |          0.15 | 0.67 |
| bandOrderReverse:band              |    -0.10 |         -0.27 |          0.08 | 0.86 |
| **conditVaried:bandOrderReverse:band** |     0.42 |          0.17 |          0.70 | 1.00 |

@@

ALM is a localist neural network model [@pageConnectionistModellingPsychology2000a], with each input node corresponding to a particular stimulus, and each output node corresponding to a particular response value. The units in the input layer activate as a function of their Gaussian similarity to the input stimulus ( a_i(X) = exp(-c(X - X_i)^2) ). So, for example, an input stimulus of value 55 would induce maximal activation of the input unit tuned to 55. Depending on the value of the generalization parameter, the nearby units (e.g., 54 and 56; 53 and 57) may also activate to some degree. The units in the input layer activate as a function of their similarity to a presented stimulus. The input layer is fully connected to the output layer, and the activation for any particular output node is simply the weighted sum of the connection weights between that node and the input activations. The network then produces a response by taking the weighted average of the output units (recall that each output unit has a value corresponding to a particular response). During training, the network receives feedback which activates each output unit as a function of its distance from the ideal level of activation necessary to produce the correct response. The connection weights between input and output units are then updated via the standard delta learning rule, where the magnitude of weight changes are controlled by a learning rate parameter.

@@

::: {.callout}

**{{< fa regular lightbulb >}} Approximate Bayesian Computation**

@@

The final component of our ABC implementation is the determination of an appropriate value of $\epsilon$. The setting of $\epsilon$ exerts strong influence on the approximated posterior distribution. Smaller values of $\epsilon$ increase the rejection rate, and improve the fidelity of the approximated posterior, while larger values result in an ABC sampler that simply reproduces the prior distribution. Because the individual participants in our dataset differed substantially in terms of the noisiness of their data, we employed an adaptive tolerance setting strategy to tailor $\epsilon$ to each individual. The initial value of $\epsilon$ was set to the overall standard deviation of each individuals velocity values. Thus, sampled parameter values that generated simulated data within a standard deviation of the observed data were accepted, while worse performing parameters were rejected. After every 300 samples the tolerance was allowed to increase only if the current acceptance rate of the algorithm was less than 1%. In such cases, the tolerance was shifted towards the average discrepancy of the 5 best samples obtained thus far. To ensure the acceptance rate did not become overly permissive, $\epsilon$ was also allowed to decrease every time a sample was accepted into the posterior.
:::

@@

::: {#tbl-task-diff}

| Dimension                                       | HTT (Project 1)                                                                                                                               | HTW (Project 2)                                                                                                |
| ----------------------------------------------- | --------------------------------------------------------------------------------------------------------------------------------------------- | -------------------------------------------------------------------------------------------------------------- |
| Task Description                                | Projectile launching to hit a target                                                                                                          | Projectile launching to hit wall at a specific velocity                                                        |
| Task Complexity                                 | More complex parabolic trajectory, both x and y velocities relevant to outcome                                                                | Simpler 1D mapping of force to outcome. Only x velocity is relevant.                                           |
| Task Space                                      | More complex: xy velocity combinations closer to the solution space may still result in worse feedback due to striking the barrier.           | Simpler: smooth, linear mapping between velocity and feedback.                                                 |
| Perceptual salience of Varied Conditions | Varied conditions (# of throwing distances) are perceptually distinct, i.e. salient differences in distance between launching box and target. | Varied conditions (# of velocity bands) are less salient - only difference is the numeral displayed on screen. |
| Testing Feedback                                | Testing always included feedback                                                                                                              | Primary testing stage had no feedback.                                                                         |
| Potential for Learning during Testing           | Limited potential for learning during testing due to feedback.                                                                                | Some potential for learning during no-feedback testing by observing ball trajectory.                           |
| Training Experience                    | Varied group gets half as much experience on any one position as the constant group.                                                          | Varied group gets 1/3 as much experience on any one velocity band as the constant group.                       |
| Testing Structure                               | Random interleaving of trained/transfer testing distances.                                                                                    | Blocked structure, separately testing trained vs extrapolation testing bands.                                  |

@@

# References
::: {#refs}
:::

>>>>

::: {.content-visible when-format="pdf"}
{{< include /Sections/pre_sections/toc_pdf.qmd >}}
:::

@@

In addition to largely overlooking the potential for non-linear generalization to confound interpretations of training manipulations, the visuomotor skill learning literature also rarely considers alternatives to schema representations [@chamberlinMemoryRepresentationMotor1992]. Although schema-theory remains influential within certain literatures, instance or exemplar-based models have accounted for human behavior across myriad domains [@jamiesonInstanceTheoryDomaingeneral2022; @loganInstanceTheoryAttention2002a]. As mentioned above, instance based accounts have been shown to perform well on a variety of different tasks with motoric components [@crumpEpisodicContributionsSequential2010; @gandolfoMotorLearningField1996a; @meighWhatMemoryRepresentation2018; @rosenbaumPlanningReachesEvaluating1995; @vandamMappingShapeVisuomotor2015]. However, such accounts have received little attention within the subdomain of visuomotor skill learning focused on the benefits of varied training.

@@

::: {#tbl-brms-models}
| Group Comparison | Code | Data |
|-----------------|---------------------------------------|-----------------|
| End of Training Accuracy | `brm(Abs. Deviation ~ condit)` | Final Training Block |
| Test Accuracy | `brm(Abs. Deviation ~ condit * bandType + (1|id) + (1|bandInt)` | All Testing trials |
| Band Discrimination | `brm(vx ~ condit * band +(1 + bandInt|id)` | All Testing Trials |

: **Statistical Model Specifications**. The specifications for the Bayesian regression models used in the analyses of each of the 3 experiments. Comparisons of accuracy use abosulte deviation as the dependent variable, while comparisons of discrimination use the raw velocities produced by participants as the dependent variable.
:::

@@

::: {#tbl-e1-train-dist}
| Term         | Estimate | 95% CrI Lower | 95% CrI Upper |  pd |
|:-------------|---------:|--------------:|--------------:|----:|
| Intercept    |   106.34 |         95.46 |        117.25 |   1 |
| conditVaried |    79.64 |         57.92 |        101.63 |   1 |

: **Experiment 1 - End of training performance**. Comparing final training block accuracy in band common to both groups. The Intercept represents the average of the baseline condition (constant training), and the conditVaried coefficient reflects the difference between the constant and varied groups. A larger positive estimates indicates a greater deviation (lower accuracy) for the varied group. CrI values indicate 95% credible intervals. pd is the probability of direction (the % of the posterior on the same side of 0 as the coefficient estimate). {.sm}
:::

@@

::: {#tbl-e1-bmm-dist}
| Term                               | Estimate | 95% CrI Lower | 95% CrI Upper |  pd |
|:-------------------|------------:|------------:|------------:|------------:|
| Intercept                          |   152.55 |         70.63 |        229.85 | 1.0 |
| conditVaried                       |    39.00 |        -21.10 |        100.81 | 0.9 |
| bandTypeExtrapolation              |    71.51 |         33.24 |        109.60 | 1.0 |
| conditVaried:bandTypeExtrapolation |    66.46 |         32.76 |         99.36 | 1.0 |

@@

::: {#tbl-e1-bmm-vx}
| Term         | Estimate | 95% CrI Lower | 95% CrI Upper |   pd |
|:-------------|---------:|--------------:|--------------:|-----:|
| Intercept    |   408.55 |        327.00 |        490.61 | 1.00 |
| conditVaried |   164.05 |         45.50 |        278.85 | 1.00 |
| Band         |     0.71 |          0.62 |          0.80 | 1.00 |
| condit\*Band |    -0.14 |         -0.26 |         -0.01 | 0.98 |

@@

::: {#tbl-e2-train-dist}
| Term         | Estimate | 95% CrI Lower | 95% CrI Upper |  pd |
|:-------------|---------:|--------------:|--------------:|----:|
| Intercept    |    91.01 |         80.67 |        101.26 |   1 |
| conditVaried |    36.15 |         16.35 |         55.67 |   1 |

@@

*Training*. @fig-e2-train-dev presents the deviations across training blocks for both constant and varied training groups. We again compared training performance on the band common to both groups (600-800). The full model results are shown in Table 1. The varied group had a significantly greater deviation than the constant group in the final training block, ( $\beta$ = `r cdtr2$estimate`, 95% CrI \[`r cdtr2$conf.low`, `r cdtr2$conf.high`\]; pd = `r cdtr2$pd`).

```{r}
modelFile <- paste0(here::here("data/model_cache/"), "e2_dist_Cond_Type_RF_2")
bmtd2 <- brm(dist ~ condit * bandType + (1|bandInt) + (1|id), 
    data=testE2, file=modelFile,
    iter=5000,chains=4, control = list(adapt_delta = .94, max_treedepth = 13))

@@

::: {#tbl-e2-bmm-dist}
| Term                               | Estimate | 95% CrI Lower | 95% CrI Upper |   pd |
|:------------------------|-----------:|-----------:|-----------:|-----------:|
| Intercept                          |   190.91 |        125.03 |        259.31 | 1.00 |
| conditVaried                       |   -20.58 |        -72.94 |         33.08 | 0.78 |
| bandTypeExtrapolation              |    38.09 |         -6.94 |         83.63 | 0.95 |
| conditVaried:bandTypeExtrapolation |    82.00 |         41.89 |        121.31 | 1.00 |

@@

::: {#tbl-e2-bmm-vx}
| Term         | Estimate | 95% CrI Lower | 95% CrI Upper |   pd |
|:-------------|---------:|--------------:|--------------:|-----:|
| Intercept    |   362.64 |        274.85 |        450.02 | 1.00 |
| conditVaried |    -8.56 |       -133.97 |        113.98 | 0.55 |
| Band         |     0.71 |          0.58 |          0.84 | 1.00 |
| condit\*Band |    -0.06 |         -0.24 |          0.13 | 0.73 |

@@

::: {#tbl-e3-train-dist}
| Term                          | Estimate | 95% CrI Lower | 95% CrI Upper |   pd |
|:----------------------|------------:|------------:|------------:|------------:|
| Intercept                     |   121.86 |        109.24 |        134.60 | 1.00 |
| conditVaried                  |    64.93 |         36.99 |         90.80 | 1.00 |
| bandOrderReverse              |     1.11 |        -16.02 |         18.16 | 0.55 |
| conditVaried:bandOrderReverse |   -77.02 |       -114.16 |        -39.61 | 1.00 |

: **Experiment 3 - End of training performance**. The Intercept represents the average of the baseline condition (constant training & original band order), the conditVaried coefficient reflects the difference between the constant and varied groups, and the bandOrderReverse coefficient reflects the difference between original and reverse order. A larger positive coefficient indicates a greater deviation (lower accuracy) for the varied group. The negative value for the interaction between condit and bandOrder indicates that varied condition with reverse order had significantly lower deviations than the varied condition with the original band order {.sm}
:::

@@

::: {#tbl-e3-bmm-dist}
| Term | Estimate | 95% CrI Lower | 95% CrI Upper | pd |
|:-----------------------------|----------:|----------:|----------:|----------:|
| Intercept | 288.65 | 199.45 | 374.07 | 1.00 |
| conditVaried | -40.19 | -104.68 | 23.13 | 0.89 |
| bandTypeExtrapolation | -23.35 | -57.28 | 10.35 | 0.92 |
| bandOrderReverse | -73.72 | -136.69 | -11.07 | 0.99 |
| **conditVaried:bandTypeExtrapolation** | 52.66 | 14.16 | 90.23 | 1.00 |
| conditVaried:bandOrderReverse | -37.48 | -123.28 | 49.37 | 0.80 |
| bandTypeExtrapolation:bandOrderReverse | 80.69 | 30.01 | 130.93 | 1.00 |
| conditVaried:bandTypeExtrapolation:bandOrder | 30.42 | -21.00 | 81.65 | 0.87 |

@@

::: {#tbl-e3-bmm-vx}
| Term | Estimate | 95% CrI Lower | 95% CrI Upper | pd |
|:-------------------------|-----------:|-----------:|-----------:|-----------:|
| Intercept | 601.83 | 504.75 | 699.42 | 1.00 |
| conditVaried | 12.18 | -134.94 | 162.78 | 0.56 |
| bandOrderReverse | 13.03 | -123.89 | 144.67 | 0.58 |
| **Band** | 0.49 | 0.36 | 0.62 | 1.00 |
| **conditVaried:bandOrderReverse** | -338.15 | -541.44 | -132.58 | 1.00 |
| conditVaried:Band | -0.04 | -0.23 | 0.15 | 0.67 |
| bandOrderReverse:band | -0.10 | -0.27 | 0.08 | 0.86 |
| **conditVaried:bandOrderReverse:band** | 0.42 | 0.17 | 0.70 | 1.00 |

@@

ALM is a localist neural network model [@pageConnectionistModellingPsychology2000a], with each input node corresponding to a particular stimulus, and each output node corresponding to a particular response value. The units in the input layer activate as a function of their Gaussian similarity to the input stimulus ( a_i(X) = exp(-c(X - X_i)\^2) ). So, for example, an input stimulus of value 55 would induce maximal activation of the input unit tuned to 55. Depending on the value of the generalization parameter, the nearby units (e.g., 54 and 56; 53 and 57) may also activate to some degree. The units in the input layer activate as a function of their similarity to a presented stimulus. The input layer is fully connected to the output layer, and the activation for any particular output node is simply the weighted sum of the connection weights between that node and the input activations. The network then produces a response by taking the weighted average of the output units (recall that each output unit has a value corresponding to a particular response). During training, the network receives feedback which activates each output unit as a function of its distance from the ideal level of activation necessary to produce the correct response. The connection weights between input and output units are then updated via the standard delta learning rule, where the magnitude of weight changes are controlled by a learning rate parameter.

@@

:::: callout
{{< fa regular lightbulb >}} **Approximate Bayesian Computation**

@@

The final component of our ABC implementation is the determination of an appropriate value of $\epsilon$. The setting of $\epsilon$ exerts strong influence on the approximated posterior distribution. Smaller values of $\epsilon$ increase the rejection rate, and improve the fidelity of the approximated posterior, while larger values result in an ABC sampler that simply reproduces the prior distribution. Because the individual participants in our dataset differed substantially in terms of the noisiness of their data, we employed an adaptive tolerance setting strategy to tailor $\epsilon$ to each individual. The initial value of $\epsilon$ was set to the overall standard deviation of each individuals velocity values. Thus, sampled parameter values that generated simulated data within a standard deviation of the observed data were accepted, while worse performing parameters were rejected. After every 300 samples the tolerance was allowed to increase only if the current acceptance rate of the algorithm was less than 1%. In such cases, the tolerance was shifted towards the average discrepancy of the 5 best samples obtained thus far. To ensure the acceptance rate did not become overly permissive, $\epsilon$ was also allowed to decrease every time a sample was accepted into the posterior.
::::

@@

::: {#tbl-task-diff}
| Dimension | HTT (Project 1) | HTW (Project 2) |
|----------------|--------------------------------|------------------------|
| Task Description | Projectile launching to hit a target | Projectile launching to hit wall at a specific velocity |
| Task Complexity | More complex parabolic trajectory, both x and y velocities relevant to outcome | Simpler 1D mapping of force to outcome. Only x velocity is relevant. |
| Task Space | More complex: xy velocity combinations closer to the solution space may still result in worse feedback due to striking the barrier. | Simpler: smooth, linear mapping between velocity and feedback. |
| Perceptual salience of Varied Conditions | Varied conditions (# of throwing distances) are perceptually distinct, i.e. salient differences in distance between launching box and target. | Varied conditions (# of velocity bands) are less salient - only difference is the numeral displayed on screen. |
| Testing Feedback | Testing always included feedback | Primary testing stage had no feedback. |
| Potential for Learning during Testing | Limited potential for learning during testing due to feedback. | Some potential for learning during no-feedback testing by observing ball trajectory. |
| Training Experience | Varied group gets half as much experience on any one position as the constant group. | Varied group gets 1/3 as much experience on any one velocity band as the constant group. |
| Testing Structure | Random interleaving of trained/transfer testing distances. | Blocked structure, separately testing trained vs extrapolation testing bands. |

@@

# References

::: {#refs}
:::
